---
title: "Data622_HW1"
author: "Daniel Sullivan"
date: "2023-03-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,include=FALSE}
library(caret)
library(tidyverse)
library(ggplot2)
library(readr)
library(corrplot)
library(gridExtra)
library(xgboost)
library(GGally)
library(e1071)
library(caTools)
library(class)
library(reshape2)
library(randomForest)
```

# First data set, from Kaggle on air quality in north Carolina

## Data exploration

begin by loading the data and modifiying date time column so it is a bit easier to work with. then get data summaries
```{r}
air_q=read.csv("https://raw.githubusercontent.com/TheSaltyCrab/Data-622/main/air-quality.csv")
air_q$Last_Check<-as.POSIXlt(air_q$Last_Check,format="%Y/%m/%d %H:%M:%S",tz=Sys.timezone())
air_q$Week_Day<-as.numeric(air_q$Last_Check$wday)
air_q$month<-as.numeric(air_q$Last_Check$mon)
air_q$hour<-as.numeric(air_q$Last_Check$hour)
air_q$Last_Check<-as.POSIXct(air_q$Last_Check,format="%Y/%m/%d %H:%M:%S",tz=Sys.timezone())

air_q<-air_q %>%
  select(!c(X,Y,Site_Label,Inside_Outside,ObjectId))
summary(air_q)
```

these below chunks show the plots of the various ppm readings/averages and targets some interesting features that were noticed. 

```{r}
#air_q$Week_Day<-air_q$Last_Check$wday
#air_q$Last_Check
#library(scales)
#head(air_q,1000)
a1=ggplot(air_q, aes(Last_Check, Current_Particulate_Matter_2_5_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

b1=ggplot(head(air_q,11000), aes(Last_Check, Current_Particulate_Matter_2_5_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
c1=ggplot(head(air_q,500), aes(Last_Check, Current_Particulate_Matter_2_5_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
d1=ggplot(tail(air_q,489), aes(Last_Check, Current_Particulate_Matter_2_5_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(a1,b1,c1,d1)
       #scale_x_datetime(labels = date_format("%H:%M:%S"))
#air_q$Last_Check[1]
```
```{r}
a2=ggplot(air_q, aes(Last_Check, PM_2_5_30_Minute_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

b2=ggplot(head(air_q,11000), aes(Last_Check, PM_2_5_30_Minute_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
c2=ggplot(head(air_q,500), aes(Last_Check, PM_2_5_30_Minute_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
d2=ggplot(tail(air_q,489), aes(Last_Check, PM_2_5_30_Minute_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(a2,b2,c2,d2)
```
```{r}
a3=ggplot(air_q, aes(Last_Check, PM_2_5_6_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

b3=ggplot(head(air_q,11000), aes(Last_Check, PM_2_5_6_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
c3=ggplot(head(air_q,500), aes(Last_Check, PM_2_5_6_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
d3=ggplot(tail(air_q,489), aes(Last_Check, PM_2_5_6_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(a3,b3,c3,d3)
```
```{r}
a4=ggplot(air_q, aes(Last_Check, PM_2_5_24_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

b4=ggplot(head(air_q,11000), aes(Last_Check, PM_2_5_24_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
c4=ggplot(head(air_q,500), aes(Last_Check, PM_2_5_24_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))
d4=ggplot(tail(air_q,489), aes(Last_Check, PM_2_5_24_Hour_Avg_)) +
       geom_point() +
       theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(a4,b4,c4,d4)
```
```{r}
a7<-ggplot(air_q, aes(x=as.character(Week_Day), y=Current_Particulate_Matter_2_5_), group=) + 
  geom_boxplot()
b7=ggplot(air_q, aes(x=as.character(month), y=Current_Particulate_Matter_2_5_), group=) + 
  geom_boxplot()
c7=ggplot(air_q, aes(x=as.character(hour), y=Current_Particulate_Matter_2_5_), group=) + 
  geom_boxplot()
grid.arrange(a7,b7,c7)

```

```{r}
#ggplot(air_q, aes(x=as.character(Week_Day), y=PM_2_5_24_Hour_Avg_)) + 
#  geom_boxplot()
a5=ggplot(air_q, aes(x=as.character(Week_Day), y=PM_2_5_24_Hour_Avg_)) + 
  geom_boxplot()
b5=ggplot(air_q, aes(x=as.character(month), y=PM_2_5_24_Hour_Avg_)) + 
  geom_boxplot()
c5=ggplot(air_q, aes(x=as.character(hour), y=PM_2_5_24_Hour_Avg_), group=) + 
  geom_boxplot()

grid.arrange(a5,b5,c5)
```
```{r}
#ggplot(air_q, aes(x=as.character(Week_Day), y=PM_2_5_24_Hour_Avg_)) + 
#  geom_boxplot()
a6=ggplot(air_q, aes(x=as.character(Week_Day), y=PM_2_5_6_Hour_Avg_)) + 
  geom_boxplot()
b6=ggplot(air_q, aes(x=as.character(month), y=PM_2_5_6_Hour_Avg_), group=) + 
  geom_boxplot()
c6=ggplot(air_q, aes(x=as.character(hour), y=PM_2_5_6_Hour_Avg_), group=) + 
  geom_boxplot()

grid.arrange(a6,b6,c6)
```

```{r}
#ggplot(air_q, aes(x=as.character(Week_Day), y=PM_2_5_24_Hour_Avg_)) + 
#  geom_boxplot()
a8=ggplot(air_q, aes(x=as.character(Week_Day), y=Temp__F_)) + 
  geom_boxplot()
b8=ggplot(air_q, aes(x=as.character(month), y=Temp__F_), group=) + 
  geom_boxplot()
c8=ggplot(air_q, aes(x=as.character(hour), y=Temp__F_), group=) + 
  geom_boxplot()

grid.arrange(a8,b8,c8)
```
correlation matrix for each data point after cleaning the data frame.
```{r}
modeling_df<-air_q %>%
  select(!c(Week_Day,hour,Latitude,Longitude,Age_of_Data_at_Check__minutes_,Last_Check,Hardware_Issues,Uptime__Seconds_,RSSI__WiFi_signal_strength_dBm_))
summary(modeling_df)
#ggpairs(modeling_df)
colnames(modeling_df)<-c('current_ppm','ppm_10min','ppm_30min','ppm_1hr','ppm_6hr','ppm_24hr','ppm1week','temp_f','humidity','pressure','month')

corrplot(cor(modeling_df))

```

## partitioning data for train and test subsets.

```{r}
set.seed(9)
x = createDataPartition(modeling_df$month, p = .8, list = F)
train_air = modeling_df[x, ]
test_air = modeling_df[-x, ]
train_air_x = data.matrix(train_air[, -11])
train_air_y = train_air[,11]
test_air_x = data.matrix(test_air[, -11])
test_air_y = test_air[, 11]
xgb_train_air = xgb.DMatrix(data = train_air_x, label = train_air_y)
xgb_test_air = xgb.DMatrix(data = test_air_x, label = test_air_y)

```

## Air Quality random forest model
```{r}
set.seed(9)
tree_model <- randomForest(
  formula = month ~ .,
  x=train_air_x,y=train_air_y, xtest = test_air_x, ytest = test_air_y
)

min<-which.min(tree_model$mse)
#train_air_x
tree_model_final <- randomForest(
  formula = month ~ .,
  data=train_air, ntree = min
)

#train_air$ppm_10min

```
```{r}
predictionT<-predict(tree_model_final, test_air_x)
#print(prediction)
predictionT<-round(predictionT,0)
#prediction1[]
#test_air$month[]
#test_air$month
cmT<-(confusionMatrix(data = factor(predictionT), reference = factor(test_air_y)))
cmT
```

## Air quality XGboost model

I modified some of the code from this article to guide me through XGboosting: https://www.statology.org/xgboost-in-r/
```{r, echo=FALSE}
#define watchlist
watchlist = list(train=xgb_train_air, test=xgb_test_air)

#fit XGBoost model and display training and testing data at each round
model = xgb.train(data = xgb_train_air, max.depth = 3, watchlist=watchlist, nrounds = 2000)

#It apears that the number of iterations that reduce test and training rmse is 1963

final_boost_1<-xgboost(data = xgb_train_air, max.depth = 3, nrounds = 1996, verbose = 0)
```

```{r}
prediction1<-predict(final_boost_1, xgb_test_air)
#print(prediction)
prediction1<-round(prediction1,0)
#prediction1<-lapply(prediction1, function(x) ifelse(x <0, 0, x))
#prediction1
for (i in 1:length(prediction1)){
  #print(i)
  if (prediction1[i]<0){
    prediction1[i]<-0
    #print(prediction1[i])
  }
  if (prediction1[i]>11){
    prediction1[i]<-11
  }
}
#test_air$month[]
#test_air$mont
#test_air_y
#unique(prediction1)
cm1<-(confusionMatrix(data = factor(prediction1), reference = factor(test_air_y)))
cm1
```



####################################################################################

# Data Set two The Pokedex

## Import second dataset the ultimate Pokedex. containing over 1000 pokemon entries. 

```{r}
pokemon_df<-read.csv("https://raw.githubusercontent.com/TheSaltyCrab/Data-622/main/pokemon.csv")

#pokemon_trim

unique(pokemon_df$type1)
length(unique(pokemon_df$type1))
```
## Data exploration

Replace miss labeled data points.and run summary statistics
```{r}
pokemon_df$type1[pokemon_df$type1=='Blastoise']<-'Water'
pokemon_df$type1[pokemon_df$type1=='Graass']<-'Grass'

pokemon_df<- pokemon_df%>%mutate(type=case_when(type1=="Grass"~1,type1=="Fire"~2,type1=="Water"~3, type1=="Bug"~4,type1=="Normal"~5,type1=="Dark"~6,type1=="Poison"~7,type1=="Electric"~8, type1=="Ground"~9,type1=="Ice"~10,type1=="Fairy"~11,type1=="Steel"~12,type1=="Fighting"~13,type1=="Psychic"~14,type1=="Rock"~15,type1=="Ghost"~16,type1=="Dragon"~17,type1=="Flying"~18))

pokemon_trim<-pokemon_df %>%
  select(!c(name,type2,number,legendary))
#unique(pokemon_df$type1)
#length(unique(pokemon_df$type1))
#summary(pokemon_trim)

summary(pokemon_trim)
#pokemon_df$type
```
plots for each stat type sorted by type
```{r}
#ggplot(air_q, aes(x=as.character(Week_Day), y=PM_2_5_24_Hour_Avg_)) + 
#  geom_boxplot()
p1=ggplot(pokemon_trim, aes(x=type1, y=total)) + 
  geom_boxplot()
p2=ggplot(pokemon_trim, aes(x=type1, y=hp)) + 
  geom_boxplot()
p3=ggplot(pokemon_trim, aes(x=type1, y=attack)) + 
  geom_boxplot()
p4=ggplot(pokemon_trim, aes(x=type1, y=defense)) + 
  geom_boxplot()
p5=ggplot(pokemon_trim, aes(x=type1, y=sp_attack)) + 
  geom_boxplot()
p6=ggplot(pokemon_trim, aes(x=type1, y=sp_defense)) + 
  geom_boxplot()
p7=ggplot(pokemon_trim, aes(x=type1, y=speed)) + 
  geom_boxplot()
p8=ggplot(pokemon_trim, aes(x=as.character(generation), y=total)) + 
  geom_boxplot()
p9=ggplot(pokemon_trim, aes(x=type1, y=generation)) + 
  geom_boxplot()
p1
p2
p3
p4
p5
p6
p7
p8
p9
grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9)
```
```{r}
pokemon_df[pokemon_trim$generation==0,]
```

```{r}

dat_m <- melt(pokemon_trim, id.vars = "type1")
#head(dat_m,50)

# plot data
fire<-dat_m[dat_m$type1=="Fire",]
water<-dat_m[dat_m$type1=="Water",]
flying<-dat_m[dat_m$type1=="Flying",]
grass<-dat_m[dat_m$type1=="Grass",]
normal<-dat_m[dat_m$type1=="Normal",]
bug<-dat_m[dat_m$type1=="Bug",]
dark<-dat_m[dat_m$type1=="Dark",]
poison<-dat_m[dat_m$type1=="Poison",]
electric<-dat_m[dat_m$type1=="Electric",]
ground<-dat_m[dat_m$type1=="Ground",]
ice<-dat_m[dat_m$type1=="Ice",]
fairy<-dat_m[dat_m$type1=="Fairy",]
steel<-dat_m[dat_m$type1=="Steel",]
fighting<-dat_m[dat_m$type1=="Fighting",]
psychic<-dat_m[dat_m$type1=="Psychic",]
rock<-dat_m[dat_m$type1=="Rock",]
ghost<-dat_m[dat_m$type1=="Ghost",]
dragon<-dat_m[dat_m$type1=="Dragon",]
fir<-ggplot(fire, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("fire")
wat<-ggplot(water, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("water")
fly<-ggplot(flying, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("flying")
gra<-ggplot(grass, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("grass")
nor<-ggplot(normal, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("normal")
g_bug<-ggplot(bug, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("bug")
dar<-ggplot(dark, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("dark")
poi<-ggplot(poison, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("poison")
ele<-ggplot(electric, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("electric")
gro<-ggplot(ground, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("ground")
g_ice<-ggplot(ice, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("ice")
fai<-ggplot(fairy, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("fairy")
ste<-ggplot(steel, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("steel")
fig<-ggplot(fighting, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("fighting")
psy<-ggplot(psychic, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("psychic")
roc<-ggplot(rock, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("rock")
gho<-ggplot(ghost, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("ghost")
dra<-ggplot(dragon, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("dragon")

grid.arrange(fir,wat,fly,gra,nor,g_bug,dar,poi,ele,gro,g_ice,fai,ste,fig,psy,roc,gho,dra)
```
plots for each type sorted by stats
```{r}
ggplot(fire, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("fire")
ggplot(water, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("water")
ggplot(flying, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("flying")
ggplot(grass, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("grass")
ggplot(normal, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("normal")
ggplot(bug, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("bug")
ggplot(dark, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("dark")
ggplot(poison, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("poison")
ggplot(electric, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("electric")
ggplot(ground, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("ground")
ggplot(ice, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("ice")
ggplot(fairy, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("fairy")
ggplot(steel, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("steel")
ggplot(fighting, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("fighting")
ggplot(psychic, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("psychic")
ggplot(rock, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("rock")
ggplot(ghost, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("ghost")
ggplot(dragon, aes(x=variable, y=value)) + 
  geom_boxplot()+coord_cartesian(ylim = c(0, 800))+ggtitle("dragon")

```
correlation matrix

```{r}
#pokemon_trim['type1']
#mon_ma <- as.matrix(pokemon_trim[, 1:9])
#mon_ma
pokemon_slim<-pokemon_trim %>%
    select(!type1)
corrplot(cor(pokemon_slim))

summary(pokemon_df$type)
```
partitioning the data into train and test sets. and scaling for kNN
```{r}
set.seed(9)
pokemon_knn<-pokemon_trim%>% select(!type1)
#head(pokemon_knn)
len<-length(pokemon_trim$type1)
#print(len)
square<-sqrt(len)
#print(square)
k<-round(square)

y = createDataPartition(pokemon_knn$type, p = .8, list = F)
train_y = pokemon_trim[y, ]
test_y = pokemon_trim[-y, ]
train_scale <- scale(train_y[, 2:9])
test_scale <- scale(test_y[, 2:9])
```

## KNN pokemon model 

explore multiple values for K
```{r}
#print(k)
for (i in 5:(k+10)){
  set.seed(9)
  classifier_knn <- knn(train = train_scale,
                      test = test_scale,
                      cl = train_y$type,
                      k=i)
  cm<-as.matrix(table(Actual = test_y$type, Predicted = classifier_knn))
  print(i)
  print(sum(diag(cm))/length(test_y$type))
  #classifier_knn
}
# best k value is 21 
```

final KNN model
```{r}
#relevel(test_y)
#levels(classifier_knn)
set.seed(9)
classifier_knn_final <- knn(train = train_scale,
                      test = test_scale,
                      cl = train_y$type,
                      k=8)
cm_knn<-(confusionMatrix(data = factor(classifier_knn_final), reference = factor(test_y$type)))

cm_knn

#cm <- confusionMatrix(test_y$type1, classifier_knn)
#cm$overall['Accuracy']
```

## Pokemon XGboost model
```{r}
set.seed(9)
#pokemon_trim<-pokemon_trim  %>% select(!type1)
p = createDataPartition(pokemon_trim$type, p = .8, list = F)
train_p = pokemon_trim[p, ]
#print(train_p$type)
test_p = pokemon_trim[-p, ]
#train_z[,10]
train_px = data.matrix(train_p[, -10])
train_py = train_p[,10]
#train_py
test_px = data.matrix(test_p[, -10])
test_py = test_p[, 10]
#length(train_zy)
#train_zx

xgb_train = xgb.DMatrix(data = train_px, label = train_py)
xgb_test = xgb.DMatrix(data = test_px, label = test_py)
#train_zy
#pokemon_trim$type
```
```{r}
#define watchlist
watchlist = list(train=xgb_train, test=xgb_test)

#fit XGBoost model and display training and testing data at each round
model = xgb.train(data = xgb_train, max.depth = 3, watchlist=watchlist, nrounds = 475)
```
#It apears that the number of iterations that reduce test and training rmse is 1963
```{r}
final_boost2<-xgboost(data = xgb_train, max.depth = 3, nrounds = 464, verbose = 0)
prediction2<-predict(final_boost2, as.matrix(test_px))
#print(prediction)
prediction2<-round(prediction2,0)
#prediction

cm2<-cm_knn<-(confusionMatrix(data = factor(prediction2), reference = factor(test_py)))
cm2
```


##########################################################################################################################

